Union and intersection contracts are hard, actually

### 2021-06-09 paper.tex
Section 3 header
Add a keyword + don't capitalise keywords
They seem uncapitalised in the acm-provided templates, so I just did
the same.
Finish reworking of 5.2

\keywords{Dynamic languages, Contracts, Union, Intersection}

@@

\unsure{TODO: CCS classification; Keywords}
\section{Introduction}
\label{sec:intro}

@@

@@

\ref{sec:issues-sem}
On the semantic side, one of the so-called the sound monitoring properties
linking the operational semantics to the denotational semantics doesn't
correspond to its attached description.
On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus obeys
all of their expected monitoring properties for higher ordered, union
and intersection contracts.
However, one of these properties states the following:

\begin{figure}[h]
$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
\caption{Sound monitoring property for intersection from Williams, Morris, and Wadler.}
\label{fig:wmw-semantics}
\end{figure}

Which means that a continuation $K$ complies negatively with an
intersection of $A$ and $B$ if it complies negatively with at least one of these.
Or, less formally, that if you have a program that is missing a value at a given point,
you can put a value wrapped with a contract stating $A \cap B$,
and expect not to get a negative violation of that contract,
if you can either put a value
wrapped with $A$, or a value wrapped with $B$.

Altough this is true, it's a weaker statement of what we'd like to have,
consider the two contexts presented on Figure \ref{fig:valid-contexts},
the first one is a valid context for a \nickel{Num -> Num} function,
that is, you could replace the $\square$ by some function wrapped in a
\nickel{Num -> Num} contract, and the context will not violate it.
Similarly, the second context from the same figure is a valid
\nickel{Bool -> Bool} context.

@@

Next, we can combine these two contexts into one, that essentially
has one single missing value, that will be used independently in both contexts
from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
This context is not a valid \nickel{Num -> Num} context, since it will apply
the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
shown on Figure \ref{fig:wmw-semantics}, it is not a valid
\nickel{Num -> Num @& Bool -> Bool} context.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (*@$\square$@*) in
(f 3, f true)
\end{lstlisting}
\caption{Invalid \nickel{Num -> Num @& Bool -> Bool} context,
combining contexts from Figure \ref{fig:valid-contexts}}
\label{fig:invalid-context}
\end{figure}

However, both our intuition, as the most common use case of intersections
(that is, overloading), tells us that this should work.
While this problem does not invalidate their work (in fact, the system they present
handles these cases correctly), it shows that existing research is still not on
good shape to handle the growing usecases being developed for
unions and intersections around dynamic languages.

>>>>

\keywords{contracts, higher-order contracts, union, intersection}

@@

\section{Introduction}
\label{sec:intro}

@@

Let us now consider union and intersection contracts, before we
explain, in Section~\ref{sec:issues-sem} how they can compromise the
properties that we described in Section~\ref{sec:feat-lang}.

@@

On the semantic side, one of the so-called the sound monitoring properties
linking the operational semantics to the denotational semantics (supposed to
capture the intuitive interpretation) is at odds with the given interpretation.
% On top of that, a fundamental problem arises on said paper,
% symptom of the difficulty of establishing good enough, and yet
% simple semantics for union and intersection dynamic checks.
% One of their main results, is the fact that their calculus obeys
% all of their expected monitoring properties for higher ordered, union
% and intersection contracts.

For a contract $A$, $\llbracket A \rrbracket^+$ defines the set of terms that
satisfy $A$. For example,

\begin{lstlisting}[language=nickel,frame=none,numbers=none]
    if true then 1 else "a" (*@$\in \llbracket Num \rrbracket^+$@*)
\end{lstlisting}

For higher-order contracts, the notion has two facets. A \nickel{Num -> Str}
contract can be violated in two ways: by the function returning a value that is
not a string, or by the caller providing an argument that is not a number.  The
former is embodied by the same notion as before: $f \in \llbracket Num \to Str
\rrbracket^+$ if whenever $f$ receives a number, it returns a string.  One
important contribution of~\cite{KeilThiemannUnionIntersection}, upon
which~\cite{RootCauseOfBlame} builds upon, is to extend the notion of contract
satisfaction to \emph{contexts} (i.e. a continuations), embodying the caller's
obligations. A context $K$ is a term with a hole to be filled with a term (see
Figure~\ref{fig:valid-contexts} for examples). Then, $K \in \llbracket Num \to
Str \rrbracket^-$ (the context $K$ satisfies the contract \nickel{Num -> Str})
if it only applies the hole to arguments satisfying the \nickel{Num} contract.

These interpretations $\llbracket . \rrbracket^+$ and $\llbracket .
\rrbracket^-$ then need to be extended to the setting of unions and
intersections. In particular, one of the definitions of~\cite{RootCauseOfBlame}
is the following:

\begin{figure}[h]
$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
\caption{Sound monitoring property for intersection from~\cite{RootCauseOfBlame}}
\label{fig:wmw-semantics}
\end{figure}

This reads as: a context $K$ satisfies the intersection of $A$ and $B$ if it
satisfies at least one of the two. Morally, the $K$s in $\llbracket A \cap B
\rrbracket^-$  should be the ones that can have their hole filled with a term
satisfying $A \cap B$ without violating the $A \cap B$ contract.

Although valid, this interpretation is weaker than what~\cite{RootCauseOfBlame}
actually describes it to be and than the semantics given
in~\cite{KeilThiemannUnionIntersection}. Consider the two contexts presented on
Figure~\ref{fig:valid-contexts}. The first one is a context satisfying
\nickel{Num -> Num}, applying the hole to a number. Similarly, the second context
from the same figure satisfies \nickel{Bool -> Bool}.

@@

Now, combining these two context as in~\ref{fig:invalid-context} gives a context
that doesn't satisfy \nickel{Num -> Num} nor \nickel{Bool -> Bool}.  According
to the semantics of~\cite{RootCauseOfBlame}, it thus doesn't satisfy \nickel{Num
-> Num @& Bool -> Bool}. Yet, our intuition tells us that this example is a
perfectly valid context for this contract: an overloaded function satisfying
both \nickel{Num -> Num} and \nickel{Bool -> Bool} can safely be applied to a
number and to a boolean in a row. And indeed, the denotational semantics given
in~\cite{KeilThiemannUnionIntersection} does accept this context as valid.

The issue is not as bad as it sounds: the system presented
in~\cite{RootCauseOfBlame} actually corresponds to the wished, stronger
semantics in practice. The specification is just not exactly the right one, but
this can probably be fixed along the lines
of~\ref{KeilThiemannUnionIntersection}, where the notion of context used to
define satisfaction is carefully restricted (\emph{elimination contexts}). While
this kind of oversight doesn't lower the quality of the work
of~\ref{RootCauseOfBlame} by any measure, it illustrates that contract
satisfaction in presence of unions, intersections and functions is a tricky
notion. This difficulty contrasts with the standard framework of higher-order
contracts where satisfaction is much more straightforward.
% Next, we can combine these two contexts into one, that essentially
% has one single missing value, that will be used independently in both contexts
% from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
% This context is not a valid \nickel{Num -> Num} context, since it will apply
% the value put on $\square$ (through \nickel{f}) to a non \nickel{Num} (\nickel{true}).
% In a similar way, it is not a valid \nickel{Bool -> Bool} context either.
% Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
% shown on Figure \ref{fig:wmw-semantics}, it is not a valid
% \nickel{Num -> Num @& Bool -> Bool} context.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (*@$\square$@*) in
(f 3, f true)
\end{lstlisting}
\caption{Combined context}
\label{fig:invalid-context}
\end{figure}

% However, both our intuition, as the most common use case of intersections
% (that is, overloading), tells us that this should work.
% While this problem does not invalidate their work (in fact, the system they present
% handles these cases correctly), it shows that existing research is still not on
% good shape to handle the growing usecases being developed for
% unions and intersections around dynamic languages.
